\chapter{Implementation}\label{chapter:implementation}

This chapter contains detailed information on the design rationale and how
solutions are implemented. It also gives instructions on how to install them. It
assumes the reader has ssh access to the server, sudo privileges, access to the
source code and has cloned the code repository on the server as described in
appendix \ref{chapter:code}.

\section{Existing Data Visualization Service}

The existing service, as detailed in~\cite{nowak2024pvw}, provides a web-based
framework for sharing and visualizing research data. The visualization
functionality leverages Kitware's ParaViewWeb Framework for browser-based remote
rendering. The ParaView-like web application allows users to directly visualize
simulation results interactively in their web browser without the need to
download large datasets. The service is realized on a LRZ Compute Cloud (CC)
instance serving ParaView Visualizer and an Apache HTTP server functioning as a
reverse proxy.

It seamlessly integrates with LRZ's Data Science Storage (DSS) as it is
connected to the CC instance via an NFS mount. Users can share specific folders
from their personal DSS directory with external researchers who do not have
direct access to the LRZ. This is achieved through unique access links, similar
to file-sharing services like Google Drive or Dropbox.

\subsection{Fixing ParaView OSMesa and glibc Issue}

In February Ubuntu 22.04 LTS upgraded its GNU C Library, commonly known as
\textit{glibc}, to version 2.35. This broke the existing data visualization
service as the used ParaView Server is incompatible with the upgraded
\textit{glibc}. The underlying problem is the \textit{OSMesa} binary used by
ParaView for software rendering and is further described in this GitLab
Issue~\cite{kitwareOsmesaBinary}. Until this report's submission date there has
not been a fix for this issue.

As proposed by the ParaView maintainers in~\cite{paraviewUsingUbuntu}
workarounds for this bug are

\begin{enumerate}
    \item Using \textit{EGL} flavor of ParaView Server
    \item Compiling ParaView with \textit{OSMesa} locally
    \item Downgrading \textit{glibc}
\end{enumerate}

Option 1 requires a CC instance with a graphics card as the \textit{EGL} variant
uses hardware rendering. However, as currently the demand for GPUs is
unquestionably high, they are no instances with GPU available. Option 2 is an
involved task with software projects of considerable size as in the case of
ParaView, making it also infeasible. Thus, we have decided on option 3 by
setting up a new CC instance running on Ubuntu 20.04 LTS that comes with
\textit{glibc} version 2.31 and installing the data visualization service again.
This is because \textit{glibc} is a very common dependency and downgrading it in
a live system would lead to further incompatibilities with other software.

\subsubsection*{Instructions for the workaround}

For the most part appendix B (Anleitung zur Bereitstellung)
of~\cite{nowak2024pvw} can be followed to install the existing service and
helper scripts on a new CC instance. Note that as the source code is not
publicly available anymore instead of using curl to download needed files, we
assume the reader already has every file needed by cloning the code repository.
After following appendix~\ref{chapter:code} to do so, the reader would need to
change to the directory that contains needed files. 

For example instead of
\begin{terminal}
    curl -fSLOJ https://raw.githubusercontent.com/FabianNowak/pv-visualizer-multi-project-config/releases/<directory>/<needed-file>
\end{terminal}

The following suffices
\begin{terminal}
    cd <directory>
\end{terminal}

The custom executable \texttt{pv-session-mapper} written in Rust has to be
compiled locally as \textit{glibc} is dynamically linked. This means at section
\textbf{pv-session-mapper installieren} follow these instructions instead

\begin{enumerate}
    \item First install Rust and its build system Cargo by using \textit{rustup}
    \begin{terminal}
        curl https://sh.rustup.rs -sSf | sh
    \end{terminal}
    It will download a script, and start the installation. If everything went
    well \texttt{Rust is installed now. Great!} should appear.
    \item Change into source directory and compile locally
    \begin{terminal}
        cd pv-session-mapper/session_mapper
        cargo build --release
    \end{terminal}
    \item Create target directory, copy and adjust custom binary
    \begin{terminal}
        sudo mkdir /opt/pv-session-mapper
        sudo cp target/release/pv-session-mapper /opt/pv-session-mapper
        sudo chown root:root /opt/pv-session-mapper/pv-session-mapper
        sudo chmod +x /opt/pv-session-mapper/pv-session-mapper
    \end{terminal}
\end{enumerate}

Once through the tutorial, these final adjustments on the Apache Server have to
be made

\begin{enumerate}
    \item Activate Apache Server module \texttt{proxy\_wstunnel} with 
    \begin{terminal}
        sudo a2enmod proxy_wstunnel
    \end{terminal}
    \item Restart Apache Server to load new configuration with
    \begin{terminal}
        sudo systemctl restart apache2
    \end{terminal}
\end{enumerate}

\subsection{Reorganizing Apache HTTP Server configuration}

The existing data visualization service will need a publicly accessible landing
page and a protected new user portal after SSO login via DFN-AAI was successful.
It is best practice to separate the web app i.e. ParaViewWeb from the marketing
site. Benefits are that non-technical staff can independently modify content of
the landing page without possibly breaking the web app. One way to achieve a
separation is with the help of subdomains. 

In our design, the landing page will be reachable under
\texttt{marge.aer.ed.tum.de} whereas the service will be bound to the subdomain
\texttt{pvw.marge.aer.ed.tum.de}. This setup also simplifies DNS configuration,
the system responsible for translating human-readable names to actual routeable
IP-addresses, as only two records are needed. An A-Record for linking
\texttt{marge.aer.ed.tum.de} to the CC instance's floating IP-address as well as
a wildcard DNS record linking \texttt{*.marge.aer.ed.tum.de} to the same
IP-address. The wildcard record will match for every subdomain, even not yet
existing ones, making this setup suitable for future requirements. 

In order to achieve aforementioned behavior, Apache's virtual host feature is
leveraged, allowing different content to be served over the same Apache server,
depending on the host name used in the request. ParaViewWeb continues to use its
existing subdirectory whereas the landing page is served from a new
subdirectory.

This means the existing configuration \texttt{paraview-multi-user.conf} is to be
moved from \texttt{conf-available} to \texttt{sites-available} directory, the
usual place where virtual configurations are placed, slightly changed to account
for the subdomain as seen in Listing~\ref{lst:002-pvw} and a new landing page
configuration shown in Listing~\ref{lst:001-marge} is to be added. Note how
Apache resolves which virtual host to serve with the help of the configured
\texttt{ServerName}.

The new landing page configuration (Listing~\ref{lst:001-marge}) resembles a
standard site configuration and is constructed in the following way:
\begin{itemize}
    \item Virtual host configurations are enclosed in a \texttt{<VirtualHost>}
    directive, meaning that everything nested within only applies to the defined
    VirtualHost.
    \item In line 2, \texttt{ServerName} specifies under which name the
    landing page is accessible.
    \item \texttt {DocumentRoot} in line 5 specifies which directory serves
    the landing page's content.
    \item \texttt{ErrorLog} and \texttt{CustomLog} define in which
    location the virtual host's error and access logs are saved.
    \texttt{APACHE\_LOG\_DIR} is a provided variable and resolves to the
    standard location \texttt{/var/log/apache2}.
    \item Lines 10-14 allow the Apache server to access the files provided in
    line 5.
\end{itemize}

\lstinputlisting[
    style=sourcecode,
    float=htp,
    label={lst:001-marge},
    caption={\texttt{001-marge.conf}: Apache configuration for Landing Page}
]{../apache-config/sites-available/001-marge.conf}

The changes made to ParaViewWeb's Apache configuration
(Listing~\ref{lst:002-pvw}) are minor and consist of enclosing the configuration
in a \texttt{<VirtualHost>} directive,  in line 2 and specifying an own
subdirectory \texttt{/var/log/apache2/pvw} for storing the web app's Apache logs
in line 12-13 to separate concerns.

\lstinputlisting[
    style=sourcecode,
    float=htp,
    label={lst:002-pvw},
    caption={\texttt{002-pvw.conf}: Apache configuration for ParaViewWeb}
]{../apache-config/sites-available/002-pvw.conf}

\subsubsection*{Instructions for reorganizing Apache HTTP Server configuration}

\begin{enumerate}
    \item Create a folder to serve files for landing page
    \texttt{marge.aer.ed.tum.de}
    \begin{terminal}
        sudo mkdir /var/www/html/marge
    \end{terminal}
    \item Create a folder to store ParaViewWeb's log files
    \begin{terminal}
        sudo mkdir /var/log/apache2/pvw
    \end{terminal}
    \item Adjust \texttt{pv-configurator} settings to reflect ParaViewWeb
    subdomain. Open and edit \texttt{configurator\_settings.json} with root
    privileges
    \begin{terminal}
        sudo vim /srv/pv-configurator/configurator_settings.json
    \end{terminal}
    \item Make following change and save file
    \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
        {
            "servername": "pvw.marge.aer.ed.tum.de",
            ...
        }
    \end{lstlisting}
    \item Change into configuration directory and copy both into Apache
    configuration
    \begin{terminal}
        cd apache-config/sites-available
        sudo cp 001-marge.conf /etc/apache2/sites-available
        sudo cp 002-pvw.conf /etc/apache2/sites-available
    \end{terminal}
    \item Deactivate old ParaViewWeb Apache configuration
    \begin{terminal}
        sudo a2disconf paraview-multi-project
    \end{terminal}
    At this point \texttt{paraview-multi-project.conf} could be deleted as it
    is no longer used.
    \item Finally, activate new configuration and restart Apache server
    \begin{terminal}
        sudo a2ensite 001-marge
        sudo a2ensite 002-pvw
        sudo systemctl restart apache2
    \end{terminal}
\end{enumerate}

\section{Creating a New User Portal}

The existing data visualization service allows users without LRZ credentials to
access shared projects via a unique access link. We want to extend the sharing
functionality to the same target group of non LRZ users. Additionally, new user
creation currently requires human involvement in processing such a request in
authenticating and creating a new Linux user on the CC instance. The goal is to
automize user management with the help of a self-service new user portal that
any authenticated user by the DFN-AAI can access. 

Our solution proposition consists of a form which is protected by DFN-AAI login,
collects necessary data and upon submission runs a server side script in the
background to create a new Linux user. Data collected include the new user's
first name, last name from which a username is generated as well as a desired
password for the new Linux account. Further, depending on whether the new user
has data on DSS to share, LRZ username (which will be the Linux username instead
of the generated) and \textit{userid} on DSS are collected. The script shall add
the user to necessary groups to allow ParaView sharing usage and set correct
\textit{userid} for DSS folder access. DSS connection has to still be done by
hand as detailed in appendix D (Anleitung: DSS-Verzeichnis mittels bindfs mit
anderen Nutzern auf der Compute-Cloud-Instanz teilen) of~\cite{nowak2024pvw}.

We realized the form with the help of an HTML form, which upon submission runs a
PHP script validating input and finally executes with sudo privileges a bash
script creating the actual Linux user. PHP has been chosen as it is well
integrated with Apache, making it a common choice for web-based applications on
Linux servers. Security is of concern as this setup allows users to execute
privileged commands on the underlying Linux server, risking potential injection
attacks. This risk will be mitigated by thoroughly sanitizing and validating
user input in the server-side PHP script. In addition, proper access control and
logging are set in place, as further described in section~\ref{section:logging}.

\subsection{HTML Form}

\lstinputlisting[
    language=html,
    style=sourcecode,
    linerange={1-2,24-48},
    float=htp,
    label={lst:htmlform},
    caption={\texttt{index.html}: New user form}
]{../newuser/index.html}

Listing~\ref{lst:htmlform} depicts the form collecting user information needed
for Linux account creation and is built the following way:

\begin{itemize}
    \item The \texttt{<form>} definition in line 5 ensures that from data will
    be sent using a \textit{POST} method and points to \texttt{create\_user.php}
    script on the server handling user creation.
    \item Line 6-13 define basic form fields for first name, last name, email
    and password. Each consists of a \texttt{label} and \texttt{input} for text,
    email and password fields respectively. The \texttt{for} attribute in the
    label is associated with the \texttt{id} of the input field to ensure
    accessibility. The \textit{name} attribute in the input field identifies
    this field with its value in the form data sent to the server. Email and
    password input types come with built-in email validation and hides entered
    characters are hidden from view.
    \item Line 15-16 is a checkbox input for users to indicate whether they want
    to share data on DSS and used to conditionally show the second set of
    fields and is further explained below.
    \item Second set of fields from line 17-22 wrapped in a \texttt{<fieldset>}
    tag are for inputting \textit{LRZ username} and \textit{uid on DSS} and also
    consist of a \texttt{label} and \texttt{input} tag. The \textit{uid} input
    is restricted to numbers bigger than 1000, the range uids are typically
    reserved for users in Linux distributions.
    \item Finally, line 24 is a submit button that sends the form data to the
    server when clicked.
\end{itemize}

\lstinputlisting[
    language=html,
    style=sourcecode,
    linerange={3-23},
    float=ht,
    label={lst:formstyle},
    caption={\texttt{index.html}: New user portal stylesheet}
]{../newuser/index.html}

The \texttt{<style>} block in \texttt{index.html} seen in
Listing~\ref{lst:formstyle} contains CSS rules for showing and hiding the input
fields for \textit{LRZ username} and \textit{uid on DSS} conditionally based on
the state of the checkbox. The rules are defined for classes \textit{control}
and \textit{conditional} and are applied respectively by assigning UI elements
with the corresponding class and works as follows:

\begin{itemize}
    \item First block specifies when class \texttt{.control} is checked then the
    \texttt{.conditional} elements that follow them in the DOM are styled to be
    visible with properties from line 3-8. 
    \item Second block specifies when class \textit{.control} is not checked
    then the \textit{.conditional} elements are styled to be hidden with
    properties from line 12-19. These properties effectively hide the
    \texttt{.conditional} elements from view while maintaining a minimal
    footprint in the
    layout.
\end{itemize}

\lstinputlisting[
    language=php,
    style=sourcecode,
    linerange={1-1, 9-43},
    float=htp,
    label={lst:phpvalidate},
    caption={\texttt{create\_user.php}: Server-side script validating user input}
]{../newuser/create_user.php}

\subsection{Create User PHP Script}

The server-side PHP script handling form submission capabilities are two-folds:
Sanitization as well as validating user input and executing create user shell
script with cleaned user input. The realized PHP script is split up in
Listing~\ref{lst:phpvalidate} and Listing~\ref{lst:phpexecute} accordingly.

Sanitization involves cleaning the input to remove or neutralize potentially
harmful characters. We need to ensure that the input does not contain characters
that could be used for command injection. Validation ensures that the input
adheres to expected formats and constraints. This step makes sure that the data
is not only sanitized but also logically correct. Listing~\ref{lst:phpvalidate}
works the following way:

\begin{itemize}
    \item \texttt{if} statement in line 2 checks if the form was submitted using
    the \textit{POST} method, ensuring that the script only processes form data
    when the form is submitted correctly.
    \item Lines 3-8 retrieve form data and assign them to variables for further
    usage and sanitize them using \texttt{escapeshellcmd}. This is a built-in
    PHP function making sure that special characters are escaped, making it safe
    to use these values in shell commands.
    \item \texttt{if} clause from line 10-13 checks if any of the required
    fields \textit{fname, lname, password} are empty. If any are missing an
    error message is displayed and the script exits.
    \item Line 15-19 use a regular expression to validate that the first name
    consists of 3 to 30 alphabetic characters, possibly followed by spaces.
    Again in case of a mismatch appropriate error message is displayed and the
    script exists.
    \item Line 20-23 is the same code as described above but validates the last
    name.
    \item Line 25-28 validates in case of a non empty \textit{lrzuser} field
    that it adheres to a LRZ username pattern (e.g., ab12xyz).
    \item If a username is provided as is the case when a user has data on DSS
    to share, line 30-35 check if the user already exists on the CC instance to
    avoid duplication. This is done by running the shell command \texttt{id -u}
    with the provided username as an argument enclosed in
    \texttt{escapeshellarg} and saving its return value in the \texttt{output}
    variable. \texttt{escapeshellarg} is a built-in PHP function escaping a
    string to be used as a single argument to a shell command. If the user
    already exists the \textit{id} command returns a non-empty value and an
    error message is shown.
\end{itemize}

\lstinputlisting[
    language=php,
    style=sourcecode,
    linerange={45-55, 58-59, 61-64},
    float=htp,
    label={lst:phpexecute},
    caption={\texttt{create\_user.php}: Server-side script executing shell script}
]{../newuser/create_user.php}

Listing~\ref{lst:phpexecute} prepares the commands to the final shell script
before it is executed:

\begin{itemize}
    \item Depending on whether \textit{lrzuser} and \textit{uid} are provided,
    line 1-9 construct the appropriate arguments for the shell command and saves
    it in the \texttt{args} variable. Arguments are constructed by concatenating
    first name, last name and optionally lrzuser or uid with their corresponding
    flags, e.g. \texttt{-f} for fname, and running them again through
    \texttt{escapeshellarg}.
    \item In line 11-12 the final command is built by specifying the shell
    script \textit{bootstrap\_user.sh} and to execute it with \textit{sudo}. The
    required password argument is only appended to the command later for logging
    reasons as described in section~\ref{section:logging}
    \item Line 13 runs the command built using \texttt{shell\_exec} while
    capturing \textit{stdout} and \textit{stderror} and echoing any possible
    output back to the user in line 15.
\end{itemize}

\subsection{Bootstrap User Shell Script}

The bootstrap user shell script starts with the usage function seen in
Listing~\ref{lst:bootstrapusage} as it's a UNIX and Linux convention to include
such in scripts and is printed if wrong arguments are provided.

\lstinputlisting[
    language=bash,
    style=sourcecode,
    linerange={3-13},
    float=htp,
    label={lst:bootstrapusage},
    caption={\texttt{bootstrap\_user.sh}: Usage function}
]{../newuser/bootstrap_user.sh}

\lstinputlisting[
    language=bash,
    style=sourcecode,
    linerange={15-36},
    float=ht,
    label={lst:bootstrapgenerate},
    caption={\texttt{bootstrap\_user.sh}: Function generating a unique username}
]{../newuser/bootstrap_user.sh}

Listing~\ref{lst:bootstrapgenerate} shows a helper function to obtain a unique
username from provided first and last name and works the following way:

\begin{itemize}
    \item First the arguments are saved in local variables \texttt{fname} and
    \texttt{lname} in line 1-2. An initial username is created by taking the
    first initial of the first name in line 5 and concatenating it with the last
    name in line 6. In line 8 the initial username is then converted to
    lowercase.
    \item In line 10-19 the loop checks if the initial username already exists
    using the \texttt{id} command. If it exists, a two decimal number (01, 02,
    etc.) is appended to the username until a unique one is found. By that
    moment the loop is exited and the found username is returned in line 21.
\end{itemize}

\lstinputlisting[
    language=bash,
    style=sourcecode,
    linerange={1-2, 37-69},
    float=htp,
    label={lst:bootstrap},
    caption={\texttt{bootstrap\_user.sh}: Create new user shell script}
]{../newuser/bootstrap_user.sh}

The actual shell script in Listing~\ref{lst:bootstrap} puts the two pieces
together:

\begin{itemize}
    \item From line 3-14 the script processes the command options by iterating
    through the arguments using a while loop. In Bash, arguments are
    automatically parsed by splitting by spaces and made available through the
    variable \texttt{\$1} for the first value, \texttt{\$2} for the second and
    so on. The condition in the loop checks if the number of remaining arguments
    is greater than zero. The \texttt{case} statement in line 4 executes either
    line 5 to 11 depending on the value of \texttt{\$1}, which is the current
    argument. Each argument is handled by matching the expected flag e.g.
    \texttt{-f} or \texttt{--fname}, assigning the following value \texttt{\$2}
    in a variable and finally using \texttt{shift} to move the argument pointer
    to the next argument, effectively skipping the current argument and its
    value. This allows the while loop to terminate after all arguments have been
    handled. Line 10 matches the case when the script is run with the help flag
    and line 11 if the current argument does not match any of the above cases.
    Both return the \textit{usage} string defined in
    Listing~\ref{lst:bootstrapusage}.
    \item In line 16-18 the \texttt{-z} test returns true if the variable is
    empty. If that is the case the \textit{usage} function is called.
    \item When the user does not need to share data on DSS and hence the
    \texttt{\$username} variable is empty, we generate a username with the help
    of the \texttt{generate\_username} function seen in
    Listing~\ref{lst:bootstrapgenerate}.
    \item The script hashes the password using the \texttt{openssl} command in
    line 21 as is required for the \texttt{useradd} command and saves it in a
    variable for later usage.
    \item In line 23-26 the shell command \texttt{useradd} used to create a new
    linux user is invoked either with or without the \texttt{\$uid} variable as
    argument depending on if the variable is empty. Other arguments provided are
    the username and first/last names which are associated with the created
    linux user. Further, every created user is assigned to the
    \texttt{pv-publisher} user group to allow immediate ParaViewWeb usage.
    \item Finally, in line 29-34 after attempting to create the user, the script
    checks the exit status of the useradd command using \texttt{\$?}, which
    stores the exit status of the last command executed. If the command a
    success message is printed. In case of failure, the outcome is echoed and
    the user is forcefully deleted with \texttt{userdel}.
\end{itemize}

As executing the \texttt{useradd} shell command requires superuser privileges,
we use \textit{sudo} to allow the \texttt{apache-proxy} user to execute only
\texttt{bootstrap\_user.sh} as root without needing a password. This is achieved
by modifying the sudoers configuration and creating a new entry of the form:

\begin{lstlisting}[frame=single]
    apache-proxy ALL=(ALL) NOPASSWD: /var/www/html/marge/newuser/bootstrap_user.sh
\end{lstlisting}


\subsection{Adding Logging}\label{section:logging}

In order to address security concerns in our new user portal, we want to put
logging in place to audit user creation. This approach enforces transparency and
comprehensiveness in case a security issue happens. In particular, we want to
detain what exact user input after sanitization is passed as arguments to the
privileged shell command. Therefore, we modify the server-side create user PHP
script with logging capabilities and log the exact command executed and its
output. A log file is created in \texttt{/var/log/newuser} and access to it
through Apache is ensured by modifying the file ownership and write access to
the \texttt{apache-proxy} user. Changes are shown in Listing~\ref{lst:logging}:

\begin{itemize}
    \item Line 2 specifies the log file's path in the variable
    \texttt{\$logfile}
    \item \texttt{log\_message} function is defined in line 4-7. It takes a
    message string and log file path string as arguments and logs the given
    message with a timestamp to the specified log file. The timestamp has the
    format year, month, day, hour, minute.
    \item To avoid logging the new user's password in plain text, we have
    constructed the variables \texttt{\$args} and \texttt{\$command} without the
    password, so that the command can be safely logged in line 9.
    \item Afterwards, the password is appended to the \texttt{\$command}
    variable, executed and finally its output also logged in line 13.
\end{itemize}

\lstinputlisting[
    language=php,
    style=sourcecode,
    linerange={1-7, 55-},
    float=htp,
    label={lst:logging},
    caption={\texttt{create\_user.php}: Server-side script logging}
]{../newuser/create_user.php}

\subsubsection*{Instructions for creating a new user portal}

\begin{enumerate}
    \item Install PHP 
    \begin{terminal}
        sudo apt install php libapache2-mod-php
    \end{terminal}
    \item Create target directory to serve the new user portal
    \begin{terminal}
        sudo mkdir /var/www/html/marge/newuser
    \end{terminal}
    \item Create log file and modify file permissions
    \begin{terminal}
        sudo mkdir /var/log/newuser
        sudo touch /var/log/newuser/create_user_php.log
        sudo chown apache-proxy:apache-proxy /var/log/newuser/create_user_php.log
        sudo chmod 640 /var/log/newuser/create_user_php.log
    \end{terminal}
    \item Change into the repository's \texttt{newuser} directory and copy HTML
    form to collect the new user's information, server-side script into target
    directory
    \begin{terminal}
        cd newuser
        sudo cp index.html /var/www/html/marge/newuser
        sudo cp create_user.php /var/www/html/marge/newuser
        sudo cp bootstrap_user.sh /var/www/html/marge/newuser
    \end{terminal}
    \item Modify file ownership and make bootstrap user shell script executable
    \begin{terminal}
        sudo chown root:root /var/www/html/marge/newuser/*
        sudo chmod +x /var/www/html/marge/newuser/bootstrap_user.sh
    \end{terminal}
    \item Create new entry to sudoers configuration by editing the file
    \begin{terminal}
        sudo visudo /etc/sudoers.d/bootstrap_user
    \end{terminal}
    \textbf{Only ever make changes with visudo, otherwise sudo can no longer be
    used in case of a syntax error.} Visudo warns if an attempt is made to save
    an incorrect file.
    \item Insert following single line and save the configuration file
    \begin{lstlisting}[frame={l}]
        apache-proxy ALL=(ALL) NOPASSWD: /var/www/html/marge/newuser/bootstrap_user.sh
    \end{lstlisting}
    This allows Apache's runtime user \texttt{apache-proxy} to execute the
    bootstrap user shell with sudo privileges.
    \item Finally, restart Apache server
    \begin{terminal}
        sudo systemctl restart apache2
    \end{terminal}
\end{enumerate}

At this point the new user portal can be accessed at
\texttt{marge.aer.ed.tum.de/newuser}.

\section{Installing and Configuring Shibboleth}\label{section:shibboleth}

We will be installing both the Shibboleth Service Provider (SP) and the
Shibboleth Embedded Discovery Service (EDS). Such an SP-side discovery service
is the method of choice in most cases, especially in scenarios where an SP
operator wants to modify or restrict the list for IdP selection via certain
filters. 

\subsection*{Instructions for Shibboleth SP}

\begin{enumerate}
    \item Install Shibboleth SP and software requirements
    \begin{terminal}
        sudo apt install libapache2-mod-shib ntp openssl
    \end{terminal}
    \item The Shibboleth SP needs an X.509 key pair for signing and encrypting
    SAML messages. Secure web servers already have a valid X.509 certificate for
    providing SSL/TLS and one could technically reuse those. However, it is
    recommended to use a dedicated self-signed certificate, independently
    configured from the SSL/TLS certificate used by the web server.
    
    Create X.509 certificate for SAML message signing and encrypting
    \begin{terminal}
        sudo shib-keygen -f -u _shibd -h marge.aer.ed.tum.de -y 3 -o /etc/shibboleth
    \end{terminal}
    \item Download DFN AAI's root certificate to the configuration directory
    \begin{terminal}
        sudo wget -O /etc/shibboleth/dfn-aai.pem https://www.aai.dfn.de/metadata/dfn-aai.pem
    \end{terminal}
    \item Shibboleth SP's configuration directory is located at
    \texttt{/etc/shibboleth}. The main configuration file is
    \texttt{shibboleth2.xml}. Each service provider needs a unique
    \textit{entityID} which we chose to be
    \texttt{https://marge.aer.ed.tum.de/shibboleth}. Currently, we are
    configuring the SP to work with the DFN-AAI test federation. Hence, the
    metadata provider is set accordingly and needs to be changed to the
    productive version once the functionality is tested. 
    
    Change into the repository's \texttt{shibboleth-config} directory and
    copy the configuration files to the target directory
    \begin{terminal}
        cd shibboleth-config
        sudo cp shibboleth2.xml /etc/shibboleth
    \end{terminal}
    \item Add following lines to the Apache configuration file
    \texttt{001-marge.conf}
    \begin{lstlisting}[frame={l}]
        UseCanonicalName On
        Include /etc/shibboleth-ds/shibboleth-ds.conf
        Redirect seeother /shibboleth https://marge.aer.ed.tum.de/Shibboleth.sso/Metadata
        RedirectMatch /start-session$ /Shibboleth.sso/Login
    \end{lstlisting}
    \item Apache and Shibboleth are well integrated through the
    \textit{mod\_shib} module. This module handles authentication and
    authorization using Shibboleth in the background by providing configurable
    Apache directives.
    
    Activate the Shibboleth module in Apache
    \begin{terminal}
        sudo a2enmod shib
    \end{terminal}
    \item After the configuration a quick test shows whether the Service
    Provider was installed properly.
    \begin{terminal}
        sudo shibd -t
    \end{terminal}
    If the last line of the output is \texttt{overall configuration is loadable,
    check console for non-fatal problems}, everything is as expected
    \item Restart Apache and the Shibboleth daemon
    \begin{terminal}
	    sudo systemctl apache2 
        sudo systemctl shibd
    \end{terminal}
    Then access \url{https://marge.aer.ed.tum.de/Shibboleth.sso/Session}. If the
    web page shows the message \textit{A valid session was not found}, the
    Shibboleth module is loaded by the Apache and is communicating with the
    \texttt{shibd} process.
    \item At this point metadata for the SP can be found at
    \url{https://marge.aer.ed.tum.de/Shibboleth.sso/Metadata}
\end{enumerate}

\subsection*{Instructions for Shibboleth EDS}

\begin{enumerate}
    \item Download and install Shibboleth EDS
    \begin{terminal}
        wget https://shibboleth.net/downloads/embedded-discovery-service/1.2.2/shibboleth-embedded-ds-1.2.2.tar.gz
        tar -xzf shibboleth-embedded-ds-1.2.2.tar.gz
        cd shibboleth-embedded-ds-1.2.2
        sudo make install
    \end{terminal}
    \item  Configure EDS by modifying the Return White List
    \begin{terminal}
        sudo vim /etc/shibboleth-ds/idpselect_config.js
    \end{terminal}
    Modify following line in the file to include the virtual host name as a
    regular expression
    \begin{lstlisting}[frame={l}]
        this.returnWhiteList = [ "^https:\/\/marge\.aer\.ed\.tum\.de\/Shibboleth\.sso\/Login.*$" ];
    \end{lstlisting}
    \item Configure backup discovery service to DFN WAYF
    \begin{terminal}
        sudo vim /etc/shibboleth-ds/index.html
    \end{terminal}
\end{enumerate}

\subsection{Access Control with Shibboleth and Apache}
One of the great advantages of Shibboleth are its capabilities to easily define
secure access control rules. A resource can be protected with access rules
defined in the web server configuration, in Shibboleth or by the application
itself. In all cases a Shibboleth session must be enforced first. This ensures
that the user's attributes are available and can be used for access control. A
user is then only granted access if his attributes match the defined access
control rules. 

Shibboleth allows using Apache directives to protect directories, files or
locations. Meaning access control rules can be defined in the virtual host's
Apache configuration files. A very simple access control rule that could be
defined in the Apache configuration looks like this: 

\begin{lstlisting}[frame=single]
    <Location /protected-directory>
        AuthType shibboleth
        ShibRequestSetting requireSession true
        Require shib-attr homeOrganizationType university
    </Location>
\end{lstlisting}

This will enforce a Shibboleth session, i.e. users have to authenticate first in
order to access the content of the directory \texttt{protected-directory} and
all its subdirectories. The three lines between the \texttt{<Location>} tag are
the Shibboleth access control directives. The \texttt{AuthType} directive
specifies the authentication method to be used. In this example, the directive
\texttt{Require shib-attr} enforces that users must be members of a university
(\texttt{homeOrganizationType university}).

\subsection{Password Protecting the New User Portal}

Until this report's submission date the DNS records for the subdomain have not
been set up due to a lack of response of the responsible administrator.
Therefore, we will not be able to test the DFN-AAI login functionality. However,
we can simulate access control instead by protecting the new user portal. This
will allow us to test the form and the server-side script without the need for
DFN-AAI login.

We employ Apache's built-in functionality to password-protect web pages with the
help of \textit{htaccess} and \textit{htpasswd} files. The \textit{htaccess}
file is a configuration file placed in the directory that needs to be protected.
Within this file, directives are set up to manage the behavior of that directory
and its subdirectories. The directives used are \textit{AuthType},
\textit{AuthName}, \textit{AuthUserFile}, and \textit{Require}: 

\begin{itemize}
	\item \textit{AuthType} is set to Basic, specifying the basic authentication
    method. 
    \item \textit{AuthName} is a descriptive label that will appear in the
    login dialog box.
    \item \textit{AuthUserFile} points to the location of the \textit{htpasswd}
    file, which contains the usernames and hashed passwords. 
    \item The \textit{Require} directive ensures that only valid users can
    access the protected directory by requiring a valid-user condition.
\end{itemize}

The \textit{htpasswd} file is where user credentials are stored. This file is
typically generated and managed using command-line tools or scripts, such as the
\texttt{htpasswd} utility that comes with Apache. Each line in the generated
\textit{htpasswd} file contains a username and a hashed version of the user's
password, separated by a colon. The hashing helps to ensure that even if the
\textit{htpasswd} file is somehow accessed by an unauthorized party, the plain
text passwords are not readily available.

When a user attempts to access a password-protected page, the browser prompts
them for a username and password. This input is then sent to the server, where
Apache verifies the credentials against the information stored in the
\textit{htpasswd} file. If the credentials match, the user is granted access;
otherwise, they are denied entry.

\subsubsection*{Instructions for password protecting the new user portal}

\begin{itemize}
    \item Create an \texttt{.htpasswd} file
    \begin{terminal}
        htpasswd -c /etc/apache2/.htpasswd admin
    \end{terminal}
    This creates a new file and adds the first user \textit{admin}. You'll be
    prompted to set a password.
    \item Create and configure \texttt{.htaccess} in the \texttt{newuser}
    directory
    \begin{terminal}
        sudo vim /var/www/html/marge/newuser/.htacess
    \end{terminal}
    Insert following lines to configure the password protection
    \begin{lstlisting}[frame={l}]
        AuthType Basic
        AuthName "Restricted Area"
        AuthUserFile /etc/apache2/.htpasswd
        Require valid-user
    \end{lstlisting}
    \item Restart Apache Server to load new configuration with
    \begin{terminal}
        sudo systemctl restart apache2
    \end{terminal} 

    When trying to access the new user portal under
    \texttt{marge.aer.ed.tum.de/newuser} the web browser prompts for a username
    and password.
    
\end{itemize}
